Query Optimization for Dynamic Imputation

### 2017-03-01 paper/algorithm.tex
paper: feedback from ben

\begin{itemize}
\item All filters are pushed to the leaves of the query tree, immediately after scanning a table. We use a pre-processing step which conjoins together filters which operate on the same table, so we can assume that each table has at most one relevant filter.

@@

\subsection{Imputation placement}
\label{sec:placement}
Imputation operators must be placed so that no relational operator encounters a tuple containing \nullv{} in an attribute that the operator examines, regardless of the state of the data in the base tables.

@@

\begin{definition}
We say that a plan $p$ with cost $\langle l, t \rangle$ dominates a plan $p'$ with cost $\langle l', t' \rangle$ if $\textsc{Dirty}(p) = \textsc{Dirty}(p') \land ((l \leq l' \land t < t') \lor (l < l' \land t \leq t'))$. We denote this as $\doms{p}{p'}$. 
\end{definition}

@@

\begin{definition}
Let $P$ be a set of plans. For $p \in P$, we say $p$ is $\alpha$-bound, if $(\textsc{Loss}(p) - \min_{p' \in P}(\textsc{Loss}(p')) \leq \alpha$. We denote this as $p^\alpha$.
We say $p$ is $\alpha$-bound optimal if $\forall p'^\alpha \in P: \textsc{Time}(p) \leq \textsc{Time}(p'^\alpha)$. 
\end{definition}

Given $\alpha$ and $P$, \ProjectName{} returns the $\alpha$-bound optimal plan in $P$. Note that the $\alpha$-bound optimal plan is unique in $P$, since $P$ contains only non-dominated plans.

@@

The computation of $\textsc{Time}(q)$ and $\textsc{Loss}(q)$ relies on having an accurate cardinality estimate of a query plan $q$ and its sub-plans. 
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
For simplicity, each of the logical nodes in a query plan points to a set of histograms.
When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for the new operation in the plan.
\Cref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.

@@

Scan costs are estimated based on the number of tuples, 
page size, and an unit-less I/O cost per page. Join costs are estimated as a function
of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
Filtering, projecting, and aggregating are all done in memory and have low computational overhead
so we assume negligible time costs for those operations. We extend these heuristics to 
account for the new operators: \textit{Drop} and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
complexity can be expressed as a function of the number of heap pages read and the I/O cost
per page.

The time computation for \textit{Impute} depends on the properties of the underlying
algorithm. \ProjectName{} treats the actual machine learning algorithm as a black box
and simply queries it for a time estimate. In the base implementation of our system,
we use an iterative algorithm (see~\Cref{sec:imputation}), where the runtime cost
is a function of the number of attributes with no missing data, number of attributes with
missing data being imputed, number of tuples, and the number of iterations for the algorithm.

In general, evaluating the time complexity for different \textit{Impute} operators is
complicated, as the properties of
different algorithms can vary significantly, the underlying I/O cost
is not easily extracted (especially in the case that the entire set of tuples does not
fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
not even consider CPU cost explicitly). For example, \cite{martin1995time} find that the
time complexity of the \textit{build tree} algorithm for one commonly-used class of
decision trees is a function of the number of classes, several overhead constants, the
parameterization of the partition and heuristic functions, and the \textit{arity} (the
number of subsets considered for each split). Indeed, the \textit{build tree} phase often
does not even dominate computation, as post-processing steps like pruning, which are vital
in achieving good performance, can have cost exponential in the height of the tree. In
practice, we find that a simple parameterization can be acceptable (i.e. yield intelligent query
plans).

@@

\begin{figure}
  \begin{align*}
    \llbracket Q[T] \rrbracket &= \begin{cases}
      P & (T, P) \in Q \\
      \emptyset & \text{otherwise} \\
    \end{cases} \\
    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned} Q \cup \{(T,\ P') \} \end{aligned}\\
                                            \text{ where } \\
                                           P' &= \left\{q ~|~ \begin{aligned}
                                                                     q \in \llbracket Q[T] \rrbracket \cup \{p\}, \\
                                                                     \textsc{Dirty}(q) = \textsc{Dirty}(p),\\
                                                                     \forall q' \in \llbracket Q[T] \rrbracket.  \doms{q} {q'} 
                                                                   \end{aligned}
                                           \right\}
  \end{align*}
  \caption{Semantics of a plan cache $Q$}
  \label{fig:semantics-plan-cache}
\end{figure}

@@

Our imputation algorithm proceeds by iteratively fitting regression trees to a subset of the
data and the target imputation columns. In each iteration, the missing values of a column are
replaced with newly imputed values. With each {\it epoch}, the quality of the imputation
improves as values progressively reflect more accurate relationships amongst attributes. The
algorithm terminates when convergence is achieved (i.e.\ the imputed values do not change
across epochs) or a fixed number of epochs is reached.

>>>>

\begin{itemize}
\item All selections are pushed to the leaves of the query tree, immediately after scanning a table. We use a pre-processing step which conjoins together filters which operate on the same table, so we can assume that each table has at most one relevant filter.

@@

\subsection{Imputation placement}
\label{sec:placement}
We place imputation operators so that no relational operator encounters a tuple containing \nullv{} in an attribute that the operator examines, regardless of the state of the data in the base tables.

@@

\begin{definition}
We say that a plan $p$ with cost $\langle l, t \rangle$ \emph{dominates} a plan $p'$ with cost $\langle l', t' \rangle$ if $\textsc{Dirty}(p) = \textsc{Dirty}(p') \land ((l \leq l' \land t < t') \lor (l < l' \land t \leq t'))$. We denote this as $\doms{p}{p'}$. 
\end{definition}

@@

\begin{definition}
Let $P$ be a set of plans. For $p \in P$, we say $p$ is $\alpha$-bound, if $(\textsc{Loss}(p) - \min_{p' \in P}(\textsc{Loss}(p')) \leq \alpha$. We denote this as $p^\alpha$.
We define the set of $\alpha$-bound plans in $P$ as $P^\alpha$.
We say a plan $p$ is $\alpha$-bound optimal if $\forall p' \in P^\alpha. \textsc{Time}(p) \leq \textsc{Time}(p'^\alpha)$. 
\end{definition}

Given $\alpha$ and $P$, \ProjectName{} returns an $\alpha$-bound optimal plan in $P$.

@@

The computation of $\textsc{Time}(q)$ and $\textsc{Loss}(q)$ relies on having an accurate cardinality estimate of a query plan $q$ and its sub-plans. 
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
For simplicity, each of the logical nodes in a query plan points to a set of histograms.
When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for the new operation in the plan, for example
redistributing the count of tuples with \nullv{} in an imputed field.
\Cref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.

@@

Scan costs are estimated based on the number of tuples, 
page size, and an unit-less I/O cost per page. Join costs are estimated as a function
of the two relations joined, any blocking operations, and repeated fetching of tuples (if required).
Filtering, projecting, and aggregating are all done in memory and have low computational overhead
so we assume negligible time costs for those operations.

We extend these heuristics to 
account for the new operators: \textit{Drop} and \textit{Impute}. \textit{Drop} is a special case of a sequential scan, and its time
complexity can be expressed as a function of the number of heap pages read and the I/O cost
per page. The time computation for \textit{Impute} depends on the properties of the underlying
algorithm. \ProjectName{} treats the actual machine learning algorithm as a black box
and simply queries it for a time estimate. In the base implementation of our system,
we use an iterative algorithm (see~\Cref{sec:imputation}), where the runtime cost
is a function of the number of attributes with no missing data, number of attributes with
missing data being imputed, number of tuples, and the number of iterations for the algorithm.

In general, evaluating the time complexity for different \textit{Impute} operators is
complicated, as the properties of
different algorithms can vary significantly, the underlying I/O cost
is not easily extracted (especially in the case that the entire set of tuples does not
fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
not even consider CPU cost explicitly). For example, \cite{martin1995time} find that the
time complexity of the \textit{build tree} algorithm for one commonly-used class of
decision trees is a function of the number of classes, several overhead constants, the
parameterization of the partition and heuristic functions, and the \textit{arity} (the
number of subsets considered for each split). The \textit{build tree} phase often
does not even dominate computation, as post-processing steps like pruning, which are vital
in achieving good performance, can have cost exponential in the height of the tree. In
practice, we find that a simple parameterization can be acceptable (i.e. yield intelligent query
plans).

@@

\begin{figure}
  \begin{align*}
    \llbracket Q[T] \rrbracket &= \begin{cases}
      P & (T, P) \in Q \\
      \emptyset & \text{otherwise} \\
    \end{cases} \\
    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned} Q \cup \{(T,\ P') \} \end{aligned}\\
                                            \text{ where } \\
                                           P' &= \left\{q ~|~ \begin{aligned}
                                                                     q \in \llbracket Q[T] \rrbracket \cup \{p\}, \\
                                                                     \textsc{Dirty}(q) = \textsc{Dirty}(p),\\
                                                                     \forall q' \in \llbracket Q[T] \rrbracket.  \doms{q} {q'} 
                                                                   \end{aligned}
                                           \right\}
  \end{align*}
  \caption{Semantics of a plan cache $Q$, where $T$ is a set of tables and $p$ is a query plan}
  \label{fig:semantics-plan-cache}
\end{figure}

@@

The algorithm used by the \textit{Impute} operator proceeds by iteratively fitting regression trees to a subset of the
data and the target imputation columns. In each iteration, the missing values of a column are
replaced with newly imputed values. With each {\it epoch}, the quality of the imputation
improves as values progressively reflect more accurate relationships amongst attributes. The
algorithm terminates when convergence is achieved (i.e.\ the imputed values do not change
across epochs) or a fixed number of epochs is reached.

### 2017-03-01 paper/conclusion.tex
paper: feedback from ben

\section{Conclusion}
Our evaluation of \ProjectName{} shows that missing values and their imputation can be successfully integrated into the relational calculus and
existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
learning technique, as operators in the algebra and use a simple, but effective, cost model to consider tradeoffs in 
imputation quality and runtime. We showed that different values for the tradeoff parameter can yield substantially
different plans, allowing the user to express their preferences for performance on a per-query basis.
Our experiments with \textit{CDC}, \textit{freeCodeCamp} and \textit{ACS} survey-based datasets
showed that \ProjectName{} can be used successfully with real-world data, expressing standard SQL queries over data with missing values. 
We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
the data exploration phase. The plans selected for each query execute an order-of-magnitude faster than
 the standard approach of imputing on the entire base tables and
then formulating queries. Furthermore, the difference in query results between the two approaches is
shown to be small in all queries considered\todo{quantify this}. Data analysts need not commit to a specific imputation strategy and can instead
vary this across queries.

We highlight the long history of dealing with missing data both in the statistical learning and database communities.
Similarly to existing work, we consider the impact of null values in databases and develop a simple set of invariants to 
successfully plan around them. In contrast to existing statistical learning work, our emphasis is not on the specific algorithm
used to impute but rather on the placement of imputation steps in the execution of a query. In contrast to existing database work,
we incorporate imputation into a cost-based optimizer and hide any details
regarding missing values inside the system, allowing users to use traditional SQL and engage in normal workloads.
While prior work has incorporated operations such as prediction into their databases, this has been domain-specific
in some cases, and in others, they have not integrated these operations into the planner. The novelty
of our contribution lies in the formalization of missing value imputation for query planning, which results in performance
gains over traditional approaches. This approach acknowledges that different users performing different queries on the same
dataset will likely have varying imputation needs and that execution plans should appropriately reflect that variation.

@@

\textbf{Imputation confidence:} Obtaining a confidence measure for query results
produced when missing values are imputed with a machine learning model is
another possible improvement to the system. In most cases, the data imputed
is unlikely to follow a standard distribution, which implies that any confidence
estimates will result from iterative approaches such as cross-validation\cite{kohavi1995study}.
In contrast to standard confidence calculations, there is the added complexity
of imputation taking place at various points in the plan and the composition of results
through standard query operations.

>>>>

\section{Conclusion}
Our evaluation of \ProjectName{} shows that missing values and their imputation can be successfully integrated into the relational calculus and
existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
learning technique, as operators in the algebra and use a simple, but effective, cost model to consider tradeoffs in 
imputation quality and runtime. We showed that different values for the tradeoff parameter can yield substantially
different plans, allowing the user to express their preferences for performance on a per-query basis.
Our experiments with \textit{CDC}, \textit{freeCodeCamp} and \textit{ACS} survey-based datasets
show that \ProjectName{} can be used successfully with real-world data, expressing standard SQL queries over data with missing values. 
We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
the data exploration phase. The plans selected for each query execute an order-of-magnitude faster than
 the standard approach of imputing on the entire base tables and
then formulating queries. Furthermore, the difference in query results between the two approaches is
shown to be small in all queries considered\todo{quantify this}. Data analysts need not commit to a specific imputation strategy and can instead
vary this across queries.

We highlight the long history of dealing with missing data both in the statistical learning and database communities.
Similarly to existing work, we consider the impact of null values in databases and develop a simple set of invariants to 
successfully plan around them. In contrast to existing statistical learning work, our emphasis is not on the specific algorithm
used to impute but rather on the placement of imputation steps in the execution of a query. In contrast to existing database work,
we incorporate imputation into a cost-based optimizer and hide any details
regarding missing values inside the system, allowing users to use traditional SQL and engage in normal workflows.
While prior work has incorporated operations such as prediction into their databases, this has been domain-specific
in some cases, and in others, they have not integrated these operations into the planner. The novelty
of our contribution lies in the formalization of missing value imputation for query planning, which results in performance
gains over traditional approaches. This approach acknowledges that different users performing different queries on the same
dataset will likely have varying imputation needs and that execution plans should appropriately reflect that variation.

@@

\textbf{Imputation confidence:} Obtaining a confidence measure for query results
produced when missing values are imputed with a machine learning model is
another possible improvement to the system. In most cases, the data imputed
is unlikely to follow a standard distribution, but iterative approaches such as cross-validation\cite{kohavi1995study}
can give confidence intervals.
In contrast to standard confidence calculations, there is the added complexity
of imputation taking place at various points in the plan, so the composition of results
through standard query operations needs to be taken into account when computing confidence intervals.

### 2017-03-01 paper/examples.tex
paper: feedback from ben

\Cref{fig:quality-plan} shows a quality-optimized plan that uses the impute operation $\mu$, which employs an iterative regression model to fill in missing values rather than dropping tuples.
It waits to impute \verb|demo.income| until after the final join has taken place, but other imputations take place earlier on in the plan, some before filtering and join operations.
Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
Meanwhile, \Cref{fig:fast-plan}'s performance-optimized plan uses the drop operation $\delta$ instead of filling in missing values.
However, it is a significant improvement over dropping all tuples with missing data, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan. In both cases, only performing
imputation on the necessary data yields a query that is much faster
than performing imputation on the whole dataset and then executing
the query.

>>>>

\Cref{fig:quality-plan} shows a quality-optimized plan that uses the impute operation $\mu$, which employs an iterative regression model to fill in missing values rather than dropping tuples.
It waits to impute \verb|demo.income| until after the final join has taken place, but other imputations take place earlier on in the plan, some before filtering and join operations.
Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
Meanwhile, \Cref{fig:fast-plan}'s performance-optimized plan uses the drop operation $\delta$ instead of filling in missing values.
However, it is a significant improvement over dropping all tuples with missing data in any field, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan. In both cases, only performing
imputation on the necessary data yields a query that is much faster
than performing imputation on the whole dataset and then executing
the query.

### 2017-03-01 paper/experiments.tex
paper: feedback from ben

\Cref{fig:runtimes} shows a summary of the performance results. We evaluated three different
configurations for each query. We compare \ProjectName{} optimizing for quality ($\alpha=0$),
\ProjectName{} optimizing for performance ($\alpha=1$), and imputing at the base tables followed
by executing the query traditionally. The quality-optimized queries are an order-of-magnitude
faster than imputing on the base tables. We can get another order-of-magnitude speedup when
optimizing for performance. This performance differential means it is feasible
to explore multiple imputations operations (including more expensive operators) when using
\ProjectName{}, in contrast to the traditional approach of base table imputation.

@@

\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)\todo{cite SMAPE} for
\ProjectName{}'s query results when compared to running imputation on the base tables and
executing the query on this cleaned base data \cite{tofallis2015better}.  Each query was
separately run 200 times in both settings and then results from the two approaches were
paired up and compared tuple-wise.  We average tuple-wise absolute percentage deviations
within each iteration of a query, and we report this value averaged over all iterations.  We
can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
to the base-imputation approach are low in all cases --- between
\lowsmapealphazero{} and \highsmapealphaoneexacs{} percent ---, indicating that on-the-fly
imputation produces similar results to imputation at the base tables.

@@

\begin{table}
\centering
\input{tables/perf_summary_13.tex}
\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$
    parameterizations relative to results when imputing on the base table. Queries optimized
    for quality ($\alpha=0$) generally achieve lower error than queries optimized for
    efficiency ($\alpha=1$). Count shares show the number of results returned from each
    query as a percentage of the number of results returned when imputing on the base table.
    A lower count share reflects more potential for errors.}
\label{table:smape}
\end{table}

\subsubsection{Runtime of Base-table Imputation}
In many real-world cases, applying the imputation step at the base table is prohibitively
expensive.
To illustrate the increasing difficulty of such an approach as datasets scale, we ran the following query over the ACS dataset:
\begin{lstlisting}[breaklines]
SELECT AVG(c0) FROM acs_dirty;
\end{lstlisting}
Applying the imputation operation to the base table is extremely expensive, in our case
completing in \acsbaseresultminutes{}. In contrast, \ProjectName{} executes a quality-optimized version
in \acsimputedbzeroresult{} and a runtime-optimized version in \acsimputedboneresult{}. This highlights the potential
benefit of using our system for early data exploration.

>>>>

\Cref{fig:runtimes} shows a summary of the performance results. We evaluated three different
configurations for each query. We compare \ProjectName{} optimizing for quality ($\alpha=0.0$),
\ProjectName{} optimizing for performance ($\alpha=1.0$), and imputing at the base tables followed
by executing the query traditionally. The quality-optimized queries are an order-of-magnitude
faster than imputing on the base tables. We can get another order-of-magnitude speedup when
optimizing for performance. This performance differential means it is feasible
to explore multiple imputations operations (including more expensive operators) when using
\ProjectName{}, in contrast to the traditional approach of base table imputation.

@@

\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE)\todo{cite SMAPE} for
\ProjectName{}'s query results when compared to the results of running imputation on the base tables and
executing the query on this cleaned base data \cite{tofallis2015better}. Given that the main datasets used
have no ground truth, this measure indicates how much error is introduced by on-the-fly imputation compared
to imputation on the base tables.

Each query was separately run 200 times in both settings and then results from the two approaches were
paired up and compared tuple-wise.  We average tuple-wise absolute percentage deviations
within each iteration of a query, and we report this value averaged over all iterations.  We
can see that optimizing for quality indeed reduces the SMAPE of query results.  In general, the SMAPE relative 
to the base-imputation approach is low in all cases --- between
\lowsmapealphazero{} and \highsmapealphaoneexacs{} percent ---, indicating that on-the-fly
imputation produces similar results to imputation at the base tables.

@@

\begin{table}
\centering
\input{tables/perf_summary_13.tex}
\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$
    parameterizations. To calculate SMAPE, \ProjectName{} results are compared to query results returned
    from executions with base table imputation. Queries optimized
    for quality ($\alpha=0$) generally achieve lower error than queries optimized for
    efficiency ($\alpha=1$). Count shares show the number of results returned from each
    query as a percentage of the number of results returned when imputing on the base table.
    A lower count share reflects more potential for errors.}
\label{table:smape}
\end{table}

\subsubsection{Runtime of Base-table Imputation}
In many real-world cases, applying the imputation step at the base table is prohibitively
expensive.
To illustrate the increasing difficulty of such an approach as datasets scale, we ran the following query over the ACS dataset:
\begin{lstlisting}[breaklines]
SELECT AVG(c0) FROM acs_dirty;
\end{lstlisting}
Applying the imputation operation to the base table is extremely expensive, as it imputes over all attributes in all rows, in our case
completing in \acsbaseresultminutes{}. In contrast, \ProjectName{} executes a quality-optimized version
in \acsimputedbzeroresult{} and a runtime-optimized version in \acsimputedboneresult{}. This highlights the potential
benefit of using our system for early data exploration. Note that the performance differential relative to base table imputation
would likely widen further if selection predicates were added to the query.

### 2017-03-01 paper/introduction.tex
paper: feedback from ben

A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.
In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. Specifically, \ProjectName{}
carefully plans the placement  imputation / row drop operations inside the query plan, resulting in a 
significant speedup in query execution while minimizing the number of dropped rows.  Also unlike previous work, the focus on of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on the  query planning involved in placement of imputation operations
in an optimal way into query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.

@@

% Scope
\ProjectName{} is designed to enable early data exploration, by allowing analysts to run their queries without explicitly decide with when to perform imputation.
% SAM -- cut the following -- could consider moving later but feels out of place here
%However, we optimize query plans with respect to a heuristic measure of imputation quality, and provide no hard guarantees about the accuracy of the results.

\subsection{Contributions}
\ProjectName{} leverages a number of contributions to minimize imputation time while producing comparable results to traditional approaches,
increasing productivity for users who frequently encounter missing values in their data. These contributions include:
\begin{itemize}
\item \textbf{Relational algebra extended with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  Impute operation fills in missing data values using an iterative regression model (CE-CART)~\cite{burgette2010multiple}.
  Drop operation simply drops tuples which contain \nullv{} values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the quality of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an additional imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
  It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
%\item \textbf{Proof of optimality:}
% We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}

>>>>

A simpler approach might drop all rows with missing values, which can not only introduce bias into the results, but also result in discarding all of the data.
In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. Specifically, \ProjectName{}
carefully plans the placement of imputation / row drop operations inside the query plan, resulting in a 
significant speedup in query execution while minimizing the number of dropped rows.  Also unlike previous work, the focus on of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on the  query planning involved in optimal placement of imputation operations
in query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.

@@

% Scope
\ProjectName{} is designed to enable early data exploration, by allowing analysts to run their queries without explicitly deciding when to perform imputation.
% SAM -- cut the following -- could consider moving later but feels out of place here
%However, we optimize query plans with respect to a heuristic measure of imputation quality, and provide no hard guarantees about the accuracy of the results.

\subsection{Contributions}
\ProjectName{} leverages a number of contributions to minimize imputation time while producing comparable results to traditional approaches,
increasing productivity for users who frequently encounter missing values in their data. These contributions include:
\begin{itemize}
\item \textbf{Relational algebra extended with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  The impute operation fills in missing data values using an iterative regression model (CE-CART)~\cite{burgette2010multiple}.
  The drop operation simply drops tuples which contain \nullv{} values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the quality of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an additional imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
  It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
%\item \textbf{Proof of optimality:}
% We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}

### 2017-03-01 paper/relatedwork.tex
paper: feedback from ben

Imputation of missing values is widely studied in the statistics and machine
learning communities. As highlighted in~\cite{gelman2006data}, missing data
can appear for a variety of reasons.
It can be missing uniformly at random and or conditioned on
existing values (observed and missing). Methods in the statistical community
focus on correctly modeling relationships between attributes to account for different forms of missingness. For example, Burgette and Reiter~\cite{burgette2010multiple} discuss the usage of sequential regression trees
for imputing missing data.

In~\cite{akande2015empirical}, Akande et al analyze the performance of various
multiple imputation techniques on the American Community Survey dataset. 
The computational difficulties of imputing on large base
tables are well known and can limit approaches. For example, Akande finds that
one approach (MI-GLM) is prohibitively expensive when attempting to impute on data that
includes variables with potentially large domains (ten categories in their case).
In contrast, \ProjectName{} allows users to specify a tradeoff between
imputation quality and runtime performance, allowing users to perform queries
directly on the entire dataset. Furthermore,
the query planner's imputation is guided by the requirements of each specific
query's operators, rather than requiring broad assumptions about query
workloads.

@@

\ProjectName's cost-based query planner 
is partially based on the seminal work developed for System R's query planning\cite{blasgen1981system}.
However, in contrast to System R, \ProjectName{} performs additional optimizations for imputing missing data and
uses histogram transformations to account for the changing nature of missing values.

\subsection{Forecasting and Databases}
Parisi et al\cite{parisi2011embedding} introduce the idea of incorporating time-series forecast operators into
databases, along with the necessary relational algebra extensions. Their work explores the theoretical
properties of forecast operators and generalizes them into a family of operators, distinguished by
the type of predictions returned. They highlight the use of forecasting for replacing missing values.
In their follow on work\cite{parisi2013temporal}, they identify various equivalence and containment
relationships when using forecast operators, which could be used to perform query plan transformations that guarantee the same result. They
explore forecast-first and forecast-last plans, which perform forecasting operations before and after executing all traditional
relational operators, respectively.

>>>>

Imputation of missing values is widely studied in the statistics and machine
learning communities. As highlighted in~\cite{gelman2006data}, missing data
can appear for a variety of reasons.
It can be missing uniformly at random or conditioned on
existing values (observed and missing). Methods in the statistical community
focus on correctly modeling relationships between attributes to account for different forms of missingness. For example, Burgette and Reiter~\cite{burgette2010multiple} discuss the usage of sequential regression trees
for imputing missing data.

~\cite{akande2015empirical} analyzes the performance of various
multiple imputation techniques on the American Community Survey dataset. 
The computational difficulties of imputing on large base
tables are well known and can limit approaches. For example, Akande finds that
one approach (MI-GLM) is prohibitively expensive when attempting to impute on data that
includes variables with potentially large domains (ten categories in their case).
In contrast, \ProjectName{} allows users to specify a tradeoff between
imputation quality and runtime performance, allowing users to perform queries
directly on the entire dataset. Furthermore,
the query planner's imputation is guided by the requirements of each specific
query's operators, rather than requiring broad assumptions about query
workloads.

@@

\ProjectName's cost-based query planner 
is partially based on the seminal work developed for System R's query planning\cite{blasgen1981system}.
However, in contrast to System R, \ProjectName{} performs additional optimizations for imputing missing data and
uses histogram transformations to account for the way relational and imputation operators affect the underlying
tuple distributions.

\subsection{Forecasting and Databases}
Parisi et al\cite{parisi2011embedding} introduce the idea of incorporating time-series forecast operators into
databases, along with the necessary relational algebra extensions. Their work explores the theoretical
properties of forecast operators and generalizes them into a family of operators, distinguished by
the type of predictions returned. They highlight the use of forecasting for replacing missing values.
In their follow on work~\cite{parisi2013temporal}, they identify various equivalence and containment
relationships when using forecast operators, which could be used to perform query plan transformations that guarantee the same result. They
explore forecast-first and forecast-last plans, which perform forecasting operations before and after executing all traditional
relational operators, respectively.
